Disparities in Climate-Induced Health Outcomes in the Greater Toronto Area.

### 2022-07-13 outputs/paper/paper.qmd
Updating paper and analysis

pollution %>%
  select(city, pollutant, mean_pollution, day, month, year) %>%
  slice(c(100, 1000, 1200, 1500, 1800, 2000, 2500, 2800, 3600, 5000, 6000, 7000, 10000, 11000, 15000)) %>%
  kable(
    caption = "Extracting rows from the Pollution data from 2002 - 2020",
    col.names=c("City", "Pollutant", "Mean Pollution", "Day", "Month", "Year"),
    linesep = "", digits = 2, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10) + theme_bw()

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = year, y = mean_pollution, factor(month_name, level = level_order))) +
  geom_point() + labs(color="Month") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2" ) %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point() + 
  xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point() +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for NO, NO2, O3 and SO2 by Month, Maximum Temperature, and Pollutant for December to March", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant != "CO" & pollutant != "SO2")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point() + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for CO by Month, Maximum Temperature, and Pollutant for December to March", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant == "CO")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point() + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for NO, NO2, O3 and SO2 by Month, Maximum Temperature, and Pollutant for April to July", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant == "SO2")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point() + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()

@@

*Health*

@@

Exposure to NO (Nitric Oxide) can irritate the lungs, reduce lung function and increase vulnerability to allergens for those with asthma.

@@

>>>>

pollution %>%
  select(city, pollutant, mean_pollution, day, month, year) %>%
  slice(c(100, 1000, 1200, 1500, 1800, 2000, 2500, 2800, 3600, 5000, 6000, 7000, 10000, 11000, 15000)) %>%
  kable(
    caption = "Extracting rows from the Pollution data from 2002 - 2020",
    col.names=c("City", "Pollutant", "Mean Pollution", "Day", "Month", "Year"),
    linesep = "", digits = 2, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10)

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = year, y = mean_pollution, factor(month_name, level = level_order))) +
  geom_point(color = "gray") + labs(color="Month") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "gray") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2" ) %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point(color = "light blue") + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "NO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "light blue") +
  ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point(color = "light pink") + 
  xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "O3") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "light pink") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point(color = "#9AD67F") + 
  labs(color="Pollutant") + xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "CO") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "#9AD67F") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = year, y = mean_pollution)) + geom_point(color = "#847FD6") + 
  xlab("Year") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

pollution %>% filter(pollutant == "SO2") %>% ggplot(
  aes(x = factor(month_name, level = level_order), y = mean_pollution)) + geom_point(color = "#847FD6") +
  xlab("Month") + ylab("Mean Pollution (ppb)") + theme_bw()
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for NO, NO2, O3 and SO2 by Maximum Temperature and Pollutant for December to March", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant != "CO" & pollutant != "SO2")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point(color = "#6DD6CD") + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()
```

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for CO by Maximum Temperature and Pollutant for December to March", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant == "CO")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point(color = "#6DD6CD") + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()
```

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.cap="Mean Pollution 2002 to 2020 for NO, NO2, O3 and SO2 by Maximum Temperature and Pollutant for April to July", fig.width=9, fig.height=3.2}

pollu_temp %>% filter((pollutant == "SO2")) %>% ggplot(aes(max_temp, mean_pollution)) + geom_point(color = "#6DD6CD") + facet_grid(. ~ pollutant) + xlab("Maximum Temperature") +  ylab("Mean Pollution") + theme_bw()

@@

NO2 is usually formed through the burning of fossil fuels, transportation services, energy production and industrial processes. Nitrogen dioxide at higher concentration levels has a strong and harsh odor and can be seen over large cities as a brownish haze. NO2 can combine with H2O and form nitric acid and nitrous acid. This falls to earth through precipitation like rain, snow, and fog, and contributed to acidification and eutrophication of ecosystems.

SO2 emissions come from natural sources like volcanoes and forest fires but these emissions are small compared to those from industrial sources. Environment Canada accounted for 1 million tonnes of man-made SO2 emissions compared to 100 tonnes of natural SO2 emissions in 2022. SO2 is seen in acid rain and due to this early recognition policies were put in place that have reduced SO2 pollution in recent decades as seen in figure 14.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=3.2}

r <- pollution %>% filter((pollutant == "NO")) %>% select(mean_pollution)
rr <- range(r)
# 0.0000 150.2917
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 13.60186

r <- pollution %>% filter((pollutant == "NO2")) %>% select(mean_pollution)
rr <- range(r)
#  0 65
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 14.97535

r <- pollution %>% filter((pollutant == "O3")) %>% select(mean_pollution)
rr <- range(r)
# 0.00000 65.95833
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 22.98821

r <- pollution %>% filter((pollutant == "SO2")) %>% select(mean_pollution)
rr <- range(r)
# -0.02500 22.79167
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 1.233524

r <- pollution %>% filter((pollutant == "CO")) %>% select(mean_pollution)
rr <- range(r)
# 0.0 1762.5
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 254.4278

pollutants <- c("NO", "NO2", "O3", "SO2", "CO")
means_pollutants <- c(13.60, 14.97, 22.99, 1.23, 254.43)

df <- data.frame(pollutants, means_pollutants)

df %>%
  select(pollutants, means_pollutants) %>%
  kable(
    caption = "Mean Pollution Levels of Each Pollutant",
    col.names=c("Pollutant", "Mean Pollution Level"),
    linesep = "", digits = 2, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10)

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=9, fig.height=3.2}

# NO

r <- pollution %>% filter((pollutant == "NO") & (year == 2002 | year == 2003 |
                                                 year == 2004 | year == 2005 | 
                                                   year == 2006)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 15.04657

r <- pollution %>% filter((pollutant == "NO") & (year == 2007 | year == 2008 |
                                                   year == 2009 | year == 2010 | 
                                                   year == 2011)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 8.573613

r <- pollution %>% filter((pollutant == "NO") & (year == 2012 | year == 2013 |
                                                   year == 2014 | year == 2015 | 
                                                   year == 2016)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 5.852171

r <- pollution %>% filter((pollutant == "NO") & (year == 2017 | year == 2018 |
                                                   year == 2019 | year == 2020)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 3.250599

# NO2

r <- pollution %>% filter((pollutant == "NO2") & (year == 2002 | year == 2003 |
                                                   year == 2004 | year == 2005 | 
                                                   year == 2006)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 19.87938

r <- pollution %>% filter((pollutant == "NO2") & (year == 2007 | year == 2008 |
                                                   year == 2009 | year == 2010 | 
                                                   year == 2011)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 15.56782

r <- pollution %>% filter((pollutant == "NO2") & (year == 2012 | year == 2013 |
                                                   year == 2014 | year == 2015 | 
                                                   year == 2016)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 13.39803

r <- pollution %>% filter((pollutant == "NO2") & (year == 2017 | year == 2018 |
                                                   year == 2019 | year == 2020)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 10.07811

# O3

r <- pollution %>% filter((pollutant == "O3") & (year == 2002 | year == 2003 |
                                                    year == 2004 | year == 2005 | 
                                                    year == 2006)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 20.94733

r <- pollution %>% filter((pollutant == "O3") & (year == 2007 | year == 2008 |
                                                    year == 2009 | year == 2010 | 
                                                    year == 2011)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 22.5484

r <- pollution %>% filter((pollutant == "O3") & (year == 2012 | year == 2013 |
                                                    year == 2014 | year == 2015 | 
                                                    year == 2016)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 23.97225

r <- pollution %>% filter((pollutant == "O3") & (year == 2017 | year == 2018 |
                                                    year == 2019 | year == 2020)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 24.85809

# SO2

r <- pollution %>% filter((pollutant == "SO2") & (year == 2002 | year == 2003 |
                                                   year == 2004 | year == 2005 | 
                                                   year == 2006)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 2.476657

r <- pollution %>% filter((pollutant == "SO2") & (year == 2007 | year == 2008 |
                                                   year == 2009 | year == 2010 | 
                                                   year == 2011)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 1.276447

r <- pollution %>% filter((pollutant == "SO2") & (year == 2012 | year == 2013 |
                                                   year == 2014 | year == 2015 | 
                                                   year == 2016)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 0.6760719

r <- pollution %>% filter((pollutant == "SO2") & (year == 2017 | year == 2018 |
                                                   year == 2019 | year == 2020)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 0.3232774

# CO

r <- pollution %>% filter((pollutant == "CO") & (year == 2002 | year == 2003 |
                                                    year == 2004 | year == 2005 | 
                                                    year == 2006)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 314.0882

r <- pollution %>% filter((pollutant == "CO") & (year == 2007 | year == 2008 |
                                                    year == 2009 | year == 2010 | 
                                                    year == 2011)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 223.6469

r <- pollution %>% filter((pollutant == "CO") & (year == 2012 | year == 2013 |
                                                    year == 2014 | year == 2015 | 
                                                    year == 2016)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 244.8328

r <- pollution %>% filter((pollutant == "CO") & (year == 2017 | year == 2018 |
                                                    year == 2019 | year == 2020)) %>% select(mean_pollution)
rm <- mean(as.numeric(r$mean_pollution),na.rm=TRUE)
# 230.332

pollutants <- c("NO", "NO", "NO", "NO", 
                "NO2", "NO2", "NO2", "NO2",
                "O3", "O3", "O3", "O3", 
                "SO2", "SO2", "SO2", "SO2", 
                "CO", "CO", "CO", "CO")
year <- c("2002 to 2006", "2007 to 2011", "2012 to 2016", "2017 to 2020", 
          "2002 to 2006", "2007 to 2011", "2012 to 2016", "2017 to 2020",
          "2002 to 2006", "2007 to 2011", "2012 to 2016", "2017 to 2020", 
          "2002 to 2006", "2007 to 2011", "2012 to 2016", "2017 to 2020", 
          "2002 to 2006", "2007 to 2011", "2012 to 2016", "2017 to 2020")
means_pollutants <- c(15.05, 8.57, 5.85, 3.25,
                      19.88, 15.57, 13.40, 10.1,
                      20.95, 22.55, 23.97, 24.86, 
                      2.48, 1.28, 0.68, 0.32, 
                      314.1, 223.65, 244.83, 230.33)

df <- data.frame(pollutants, means_pollutants, year)
df %>%
  select(pollutants, means_pollutants, year) %>%
  kable(
    caption = "Mean Pollution Levels of Each Pollutant by 5-Year Intervals",
    col.names=c("Pollutant", "Mean Pollution Level", "5-Year Interval"),
    linesep = "", digits = 2, booktabs=TRUE) %>%
  kable_styling(latex_options = "HOLD_position", font_size = 10)
```

@@

Exposure to NO (Nitric Oxide) can irritate the lungs, reduce lung function and increase vulnerability to allergens for those with asthma. Nitric Oxide is seen in smog and can react with other substances to harm human health.

@@

SO2 (Sulfur Dioxide) can harm human respiratory systems by triggering respiratory illness and worsening existing respiratory and cardiovascular diseases. Sulfur dioxide is seen in smog and can react with other substances like nitric oxide to harm human health. When SO2 transforms into sulfate particles and combines with other particles in the atmosphere it can contribute to the formation of particulate matter (PM2.5) which harms human health as well as the environment. Sensitive populations such as children, the elderly, those with asthma are more susceptible to the affects of SO2.

NO2 (Nitrogen Dioxide) has consequences on human health including emergency hospitalizations for respiratory conditions for adults that live in lower income communities in Toronto. Children are more vulnerable to these effects of respiratory damage. Children that have been exposed to NO2 have an increased risk of asthma, wheezing, infections in the ear, nose and throat, influenza, serious colds, and reduced lung function.
